---
title: "Residual plots for linear models using the nullabor package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Residual plots using the nullabor package}
  %\VignetteEncoding{UTF-8}
---  


Residual plots using the **nullabor** packlage
=======================================


```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center")
```

The  **nullabor** package provides functions to draw residual plots for linear regression models using the lineup package.

```{r message=FALSE}
library(nullabor)
```

First, fit a linear model:
```{r message=FALSE}
data(tips)
x <- lm(tip ~ total_bill, data = tips)
```

The `lineup_residuals` function can now be used to generate four types of residual lineup plots.

The first residual plot shows the residuals versus the fitted values. It is used to test the hypothesis that the response variable is a linear combination of the predictors:

```{r message=FALSE}
lineup_residuals(x, type = 1)
```

The second plot is a normal Q-Q plot for the residuals, used to test the hypothesis that the errors are normal:

```{r message=FALSE}
lineup_residuals(x, type = 2)
```

The third plot is a scale-location plot used to test the hypothesis that the errors are homoscedastic:

```{r message=FALSE}
lineup_residuals(x, type = 3)
```

The fourth plot shows leverage, and is used to identify points with high residuals and high leverage, which are likely to have a strong influence on the model fit:

```{r message=FALSE}
lineup_residuals(x, type = 4)
```

The plots are created using `ggplot2` and can be modified in the same way as other ggplots. In addition, `lineup_residuals` has arguments for changing the colors used:

```{r message=FALSE}
library(ggplot2)
lineup_residuals(x, type = 3,
                color_points = "skyblue",
                color_trends = "darkorange") +
    theme_minimal()
```

If the null hypothesis in the type 1 plot is violated, consider using a different model. If the null hypotheses in the type 2 or 3 plots
are violated, consider using bootstrap p-values; see [Section 8.1.5](https://www.modernstatisticswithr.com/regression.html#bootreg1) of Thulin (2024) for details and recommendations.


References
----------

Buja, A., Cook, D., Hofmann, H., Lawrence, M., Lee, E.-K., Swayne, D. F, Wickham, H. (2009) Statistical Inference for Exploratory Data Analysis and Model Diagnostics, Royal Society Philosophical Transactions A, 367:4361--4383, DOI: 10.1098/rsta.2009.0120.

Thulin, M. (2024) _Modern Statistics with R_. Boca Raton: CRC Press. ISBN 9781032512440. [https://www.modernstatisticswithr.com/](https://www.modernstatisticswithr.com/)
